---
title: "Working with cyclic trajectories"
author: "Kaitlyn Ramesh"
affiliation: "Northeastern University"
date: "April 25, 2025"
output: 
  rmarkdown::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Working with cyclic trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Overview
In this vignette, we will discuss how RETRO can help infer pseudotime for single-cell RNA-seq datasets with oscillating gene expression patterns. Due to its use of sampling time, trajectory inference with RETRO correctly orders cells3 that return to an earlier state.

## 2. Analysis
```{r}
library(RETRO)

# Additional packages for vignette 
library(Biobase) # data storage + processing
library(ggplot2) # plotting
library(doParallel)
library(parallel)
```

### 2.1 Preprocessed data
```{r}
cyclic_url = "https://raw.githubusercontent.com/kaitlynramesh/RETRO-data/main/synthetic/scd_twocycles.rds"
scd_twocycles = readRDS(gzcon(url(cyclic_url)))
```

### 2.2 Dimensionality reduction
```{r}
# PCA analysis
pca = experimentData(scd_twocycles)@other$PCA

time = phenoData(scd_twocycles)@data[["time"]] 
time_label = as.factor(time) # factor for coloring
ggplot(as.data.frame(pca$x), aes(x=PC1,y=PC2,colour=time_label)) + geom_point() + theme_bw()
```

## 3. RETRO

### 3.1 Path detection
- same weight coord approach
- need to cluster indiscriminately to cover entire projection
- lambda = 1 will incorporate time information for path detection
#### 3.1.1 Integrating gene expression and time information in distance
```{r}
lambda = 1

coordinates <- weight_coord(scd_twocycles, weight = lambda)
coordinates_uw <- weight_coord(scd_twocycles, weight = 0)
```

#### 3.1.2 Re-clustering to separate cycles
- specify period!!!!
```{r}
k_range = seq(9,14) # set range of clusters 
num_scores = 20
period = 4

retro_obj = set_RETRO_class(coordinates=coordinates,
                            coordinates_uw=coordinates_uw,
                            k_range=k_range,
                            time=time,
                            period=period)
```

#### 3.1.3 Iteratively scoring MSTs
```{r}
if (!interactive()) {
  foreach::registerDoSEQ()
} else {
  ncores = parallel::detectCores()
  registerDoParallel(cores = ncores)
}
retro_obj = scoring(retro_obj, k_range, num_scores)

boxplot_scoring(retro_obj) # visualization distribution of score over K clusters
```

#### 3.1.4 Determine consensus number of lineages
- should be 1 lineage, so not heatmap is plotted (all cells map to 1 lineage)
```{r}
retro_obj <- get_num_lineages(retro_obj, percent=0.1) 
```

### 3.2 Pseudotime estimation 
#### 3.2.1 Spline fitting
```{r}
retro_obj <- get_bezier_curve(retro_obj)
retro_obj <- get_mapped_cells(retro_obj) # map any left over cells to lineage
```

#### 3.2.2 Fitting real time-calibrated pseudotime
```{r}
retro_obj = pseudotime_fit(retro_obj)

cyclic_lineage = retro_obj@pseudotime_graph[[1]]
cyclic_lineage
```

### 3.3 Pseudotime visualization
- can visualize inferred pseudotime like normal 
#### 3.3.1 Heatmap
```{r}
coordinates = retro_obj@coordinates
pseudotime = retro_obj@pseudotime
curve = retro_obj@RETRO_Curve[[1]][["Full_Curve"]] # curve data points

ggplot(as.data.frame(coordinates), aes(x=PC1, y=PC2, colour=pseudotime)) +
  geom_point() +
  scale_colour_continuous(low="blue", high="orange") +
  geom_point(data = as.data.frame(curve[,1:2]), aes(x = V1, y = V2), size=0.5, colour='black') +
  theme_bw()
```

#### 3.3.2 Density plot
```{r}
pseudotime_density(time, pseudotime)
```


## 4. Session info
```{r}
sessionInfo()
```

