---
title: "An introduction to RETRO"
author: "Kaitlyn Ramesh"
affiliation: "Northeastern University"
date: "April 25, 2025"
output: 
  rmarkdown::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{An introduction to RETRO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Overview
In this vignette, we will discuss how to use the package RETRO to infer the pseudotime for simulated time-series scRNA-seq data. RETRO provides a workflow for incorporating sampling time in trajectory inference such that the outputted pseudotime is consistent with the dynamics of cellular transitions. We will also show how RETRO allows for capturing both branching and cyclic trajectories.

## 2. Analysis
```{r}
library(RETRO)
```
We will make use of a few packages for plotting (`ggpubr`, `ggplot2`) to aid in visualization, as well as packages for parallelization (`doParallel`, `parallel`) to expedite steps in RETRO.
```{r}
library(ggplot2) # plotting
library(ggpubr) # plotting
library(doParallel)
library(parallel)
```

### 2.1 Preprocessed data
```{r}
# Example 2: Bifurcating lineages
data("scd_bifurcation")
names(scd_bifurcation) # (1) log-normalized count matrix, (2) discrete sampling time, (3) ground-truth continuous developmental time 

# Example 1: Single lineage, two cycles
data("scd_twocycles")

# Time information (example 1 bifurcation)
time = scd_bifurcation$Time
print(sort(unique(time)))
```
We will be looking at two examples of time-series scRNA-seq in this tutorial, both simulated cases generated using the `dyntoy` library from dynverse. The first example is a bifurcation where a starting population of cells splits into two terminal cell types. We contain log-transformed counts in `scd_bifurcation[["Counts]]`, discrete experimental time in `scd_bifurcation[["Time"]]`, and ground truth continuous time in `scd_bifurcation[["Time_Cont"]]`.
<br>
The second example forms two cycles, showing that the cells return to earlier states of gene expression at later time points. Because `dyntoy::generate_dataset()` can only form a single cycle, we modified its simulation to become two overlapping cycles. As a result, there is no ground-truth developmental time for this case. 

### 2.2 Dimensionality reduction
```{r}
# PCA analysis
pca = prcomp(scd_bifurcation$Counts, center=T)

time_label = as.factor(time) # factor for coloring
ggplot(as.data.frame(pca$x), aes(x=PC1,y=PC2,colour=time_label)) + geom_point() + theme_bw()
```

For this first example, we will use PCA for dimensionality reduction. We recommend linear methods for dimensionality reduction although RETRO is compatible with UMAP coordinates as an input. 

## 3. RETRO
RETRO path detection starts by concatenating weighted time information to the coordinates determined using dimensionality reduction methods. There are two versions of coordinates used. If the parameter `weight > 0`, then there is a equal contribution from time and gene expression in this concatenated time vector. We can increase `weight` to increase contribution from the sampling time. We use `coordinates_uw` where `weight = 0` for clustering since on a matrix of coordinates without time information, we can ensure that the initial set of MST nodes covers the entire set of coordinates. 
<br>
The parameter range `k_range` is important for generating a wide variety of clustering results. We recommend that the range starts around the number of time-points in the data and includes at least 6 values for MST generation.  

### 3.1 Path detection
#### 3.1.1 Integrating gene expression and time information in distance
```{r}
lambda = 1
time <- scd_bifurcation$Time

scd_bifurcation$PCA = pca
coordinates <- weight_coord(scd_bifurcation, weight = lambda)
coordinates_uw <- weight_coord(scd_bifurcation, weight = 0)
```

#### 3.1.2 Iterative clustering to identify global lineages
We start off analysis with RETRO by creating `retro_obj`, which stores the coordinate information, sampling time. We also set the desired cluster range 
`k_range`. For each K, we iteratively cluster the cells a number of times set by the parameter `num_scores`. By default it is ~100, but for the purposes of this vignette, we will limit the runs to 20. 
<br>
Using `set_RETRO_class()`, we establish an RETRO object `retro_obj`, which is important for storing parameters, inputs, intermediates, and outputs all involved in conducting pseudotime analysis with RETRO.
```{r}
k_range = seq(8,15) # set range of clusters 
num_scores = 20

retro_obj = set_RETRO_class(coordinates=coordinates,
                            coordinates_uw=coordinates_uw,
                            k_range = k_range,
                            time=time)

ncores <- parallel::detectCores()
registerDoParallel(cores=ncores)
retro_obj = scoring(retro_obj, k_range, num_scores)
```

#### 3.1.3 Visualize distribution of scores over K clusters
Here, we can visualize results from the MST generation and scoring steps. We can see that with more clusters, the score decreases, indicating that the MSTs fit closer along the projection of cells. This makes sense as for this bifurcating structure, more clusters will increase the resolution of the MST structure. 
```{r}
boxplot_scoring(retro_obj)
```

#### 3.1.4 Identify global lineages across the ensemble of MSTs
`get_num_lineages()` is important for inferring the global trajectories underlying the data based on results of the iterative clustering and MST generation steps. The parameter `percent` is default at 0.05 to select the top 5% of high performing MSTs to determine the optimal number of lineages, as well as group cells based on their lineage. Because `num_scores` is low, we will select the top 10% of MSTs (`percent = 0.1`) for analysis.
```{r}
retro_obj <- get_num_lineages(retro_obj, percent=0.1) 
```


### 3.2 Pseudotime estimation 
#### 3.2.1 Spline fitting
```{r}
retro_obj <- get_bezier_curve(retro_obj)
retro_obj <- get_mapped_cells(retro_obj)
```

#### 3.2.2 Fitting real time-calibrated pseudotime
```{r}
retro_obj = pseudotime_fit(retro_obj)

lin_1 = retro_obj@pseudotime_graph[[1]]
lin_2 = retro_obj@pseudotime_graph[[2]]

ggarrange(lin_1, lin_2, ncol=2)
```

### 3.3 Pseudotime visualization
#### 3.3.1 Heatmap
```{r}
coordinates = retro_obj@coordinates
pseudotime = retro_obj@pseudotime
curves = sapply(retro_obj@RETRO_Curve, "[", 3)

ggplot(as.data.frame(coordinates), aes(x=PC1, y=PC2, colour=pseudotime)) +
  geom_point() +
  scale_colour_continuous(low="blue", high="orange") +
  geom_point(data = as.data.frame(curves[[1]][,1:2]), aes(x = V1, y = V2), size=0.5, colour='black') +
  geom_point(data = as.data.frame(curves[[2]][,1:2]), aes(x = V1, y = V2), size=0.5, colour='black') +
  theme_bw()
```

#### 3.3.2 Density plot
```{r}
psupertime_density(time, pseudotime)
```


## 4. Session info
```{r}
sessionInfo()
```

